# Analysis of Longitudinal Data

Loading the data

```{r}
library(tidyverse)
RatsL <- read_csv("data/RATSL.csv")
bprsL <- read_csv("data/BPRSL.csv")

RatsL %>% str
bprsL %>% str

RatsL %>% summary
bprsL %>% summary
```

There are two datasets for this week's analysis. Both have been converted into long format, hence the *L* in their names. The *RatsL* data has 16 rats (ID) divided into 3 groups (Group) and the weight (weight) of each was measured several times during 64 days (Time).

In the *bprsL* dataset , short for brief psychiatric rating scale, a test score (bprs) was measured weekly (week) in 20 subjects (subject) per two treatment groups (treatment).

From the summaries we can see that the both datasets include quite a lot of variation.

Transforming categorical variables from numerical into factors:

```{r}
RatsL$ID <- factor(RatsL$ID)
RatsL$Group <- factor (RatsL$Group)

bprsL$treatment <- factor(bprsL$treatment)
bprsL$subject <- factor(bprsL$subject)

RatsL %>% str
bprsL %>% str
```

Great, now we can start diving in.

### RatsL

First, plotting RatsL data:

```{r}
library(ggplot2)
ggplot(RatsL, aes(x = Time, y = weight, group = ID)) +
  geom_line(aes(col = Group)) +
  theme(legend.position = "top") +
  scale_x_continuous(name = "Time (days)", 
                     breaks = seq(0, 60, 10)) +
  scale_y_continuous(name = "Weight (grams)",
                     limits = c(min(RatsL$weight), max(RatsL$weight)))
```

There seems to be an increasing trend in weight in each group, but the groups have notably different starting weights. Group one seems to increase in weight less than groups 2 and 3.

Next, standardizing the data and plotting it:

```{r}
RatsL <- RatsL %>% 
  group_by(Time) %>% 
  mutate(stdprs = scale(weight)) %>% 
  ungroup

ggplot(RatsL, aes(x = Time, y = stdprs, group = ID)) +
  geom_line(aes(col = Group)) +
  theme(legend.position = "top") +
  scale_x_continuous(name = "Time (days)", 
                     breaks = seq(0, 60, 10)) +
  scale_y_continuous(name = "Standardized Weight")
```

Honestly, I don't know what this adds that we couldn't see from the previous plot.

Moving on to a summary graph:

```{r}
n <- 16 # sample size

# creating a summary dataset
RatsLS <- RatsL %>% 
  group_by(Group, Time) %>% 
  summarise(mean = mean(weight), se = (sd(weight)/sqrt(n))) %>% 
  ungroup

# plotting the summary
RatsLS %>% ggplot(
  aes(x = Time, y = mean, shape = Group, col = Group)) +
  geom_line() +
  theme(legend.position = "top") +
  scale_x_continuous(name = "Time (days)", 
                     breaks = seq(0, 60, 10)) +
   geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=1) +
  geom_point(size = 1) +
  scale_y_continuous(name = "mean(weight) +/- se(weight)")
```

The summary graph shows us group averages and their standard errors instead of individuals. This makes the graph the easier to read the bigger the data.

Next, checking for outliers in the data using box plots:

```{r}
outlaw_rat <- RatsL %>% 
  filter(Time > 1) %>% # removing baseline
  group_by(ID, Group) %>% 
  summarise(mean = mean(weight)) %>% 
  ungroup

outlaw_rat %>% ggplot(
  aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape = 23, size = 2, fill = "pink") +
  scale_y_continuous(name = "mean(weight, days 1-64")
```

Seems that each group has an outlier. I'm dubious of removing anything from such a small sample, but I'm gonna do it anyway for practice and replot.

```{r}
outlaw_rat1 <- outlaw_rat %>% filter(mean > 240, # group 1 outlier
                                     mean < 550, # group 2 outlier
                                     mean > 500 | Group != 3) 
                                                 # group 3 outlier
# plotting
outlaw_rat1 %>% ggplot(
  aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape = 23, size = 2, fill = "pink") +
  scale_y_continuous(name = "mean(weight, days 1-64")
```

Yay, done. Now, it's obvious enough just from the graphs that these groups of rats have very different weight, but let's test that quantitatively anyway for the fun of it. Performing a pairwise t-test:

```{r}
library(rstatix)
outlaw_rat1 %>%
  pairwise_t_test(mean ~ Group, p.adjust.method = "bonferroni")
```

As the test shows, the groups have different weight at amusingly low p-values.

Finally, calculating a linear model for the data:

```{r}
# loading the starting weight of the rats
baseline <- RatsL %>% filter(Time==1) %>% select(weight) %>% as.vector

# adding the baseline weight as a column to a data frame
outlaw_rat2 <- outlaw_rat %>% 
  mutate(baseline = baseline$weight)

# calculating a linear model for the data
rat_fit <- lm(mean ~ baseline + Group, data = outlaw_rat2)

rat_fit %>% anova
```

The ANOVA table shows us that the baseline weight is strongly related to the later weight measurements, whereas with a p-value of about 0.076 we cannot be very confident in saying that the groups had a difference in weight gain.

### bprsL

A reminder of the variables in the data:

```{r}
str(bprsL)
```

Okay, time to plot it:

```{r}
bprsL %>% ggplot(
  aes(x = week, y = bprs, col = subject)) +
  geom_line() +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme(legend.position = "none") +
  scale_x_continuous(breaks = seq(0, 8, 1))
```

Both treatment groups seem to have a linear reduction in bprs scores during the eight week treatment period. I wonder what a placebo group would look like.

Fitting a linear model:

```{r}
bprs_reg <- lm(bprs ~ week + treatment, bprsL)
bprs_reg %>% summary
```

Fitting a random intercept model:

```{r}
library(lme4)
bprs_ref <- lmer(bprs ~ week + treatment + (1 | subject), bprsL, REML = FALSE)
bprs_ref
```

With two random intercepts:

```{r}
bprs_ref1 <- lmer(bprs ~ week + treatment + (week | subject), bprsL, REML = FALSE)
bprs_ref1 %>% summary

anova(bprs_ref1, bprs_ref)
```

The standard error for treatment was lower in this linear mixed model compared to the linear regression model.

Trying *week + treatment* to *week \* treatment*

```{r}

bprs_ref2 <- lmer(bprs ~ week * treatment + (week | subject), bprsL, REML = FALSE)
bprs_ref2 %>% summary

anova(bprs_ref2, bprs_ref1)
```

In this one the standard error was again larger. None of these model show evidence that there would be a difference in bprs score between the treatment groups.

Plotting the model with the smallest standard error:

```{r}
bprsL$fitted <- fitted(bprs_ref1) 

bprsL %>% ggplot(aes(x = week, y = fitted, group = subject)) +
  geom_line(aes(col = subject)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme(legend.position = "none") +
  scale_x_continuous(breaks = seq(0, 8, 1))
```

As can be seen from these fitted plots, it's hard to even see difference between them (I'm not lying there really is some if you look suuuper closely).
