# Dimensionality Reduction Techniques

## Human data

> Show a graphical overview of the data and show summaries of the variables in the data. Describe and interpret the outputs, commenting on the distributions of the variables and the relationships between them. (0-3 points)

```{r}
library(tidyverse)
human <- read_csv("data/human.csv")
human %>% summary
human %>% str
```

The data contains 8 variables and 155 observations.

-   eduFM: ratio of females to males with secondary education

-   workFM: female to male ratio of labor force participation

-   exp.edu: the expected years of education

-   life.exp: life expectancy at birth in years

-   momdeath: maternal mortality ratio

-   teenmom: adolescent birth rate

-   repre: percent representation in parlament

```{r}
library(GGally)
human %>% ggpairs
```

```{r}
library(corrplot)
cor(human) %>% corrplot(title = "Correlations")
```

### PCA on non-standardized data

> Perform principal component analysis (PCA) on the raw (non-standardized) human data. Show the variability captured by the principal components. Draw a biplot displaying the observations by the first two principal components (PC1 coordinate in x-axis, PC2 coordinate in y-axis), along with arrows representing the original variables. (0-2 points)

```{r}
pca <- prcomp(human)
pca

pca %>% biplot(cex = c(0.7, 1))
```

### PCA on standardized data

> Standardize the variables in the human data and repeat the above analysis. Interpret the results of both analysis (with and without standardizing). Are the results different? Why or why not? Include captions (brief descriptions) in your plots where you describe the results by using not just your variable names, but the actual phenomena they relate to. (0-4 points)

```{r}
human_std <- scale(human) # standardizing the data
pca_std <- prcomp(human_std) # performing pca

# creating axis labels
s <- pca_std %>% summary
perc <- round(100*s$importance[2, ], digits = 2)
pc_lab <- paste0(names(perc), " (", perc, "%)")

# plotting 
pca_std %>% biplot(
  main = "PCA on standardized data", 
  cex = c(0.7, 1), 
  col = c("black","red"),
  xlab = pc_lab[1], 
  ylab = pc_lab[2],
  sub = "subtext")
```

The results are very different.

> Give your personal interpretations of the first two principal component dimensions based on the biplot drawn after PCA on the standardized human data. (0-2 points)

The PC1 explains 53.61 % of the variance and captures well the educational and health-related variables. The PC2 explains 16.24 % of the variance and captures the variables repre and workFM. We can see the same relationships in the correlationplot: the health and educational variables correlate a lot with each other, whereas *repre* and *workFM* do not.

## Tea data

> Load the tea dataset from the text file <https://raw.githubusercontent.com/KimmoVehkalahti/Helsinki-Open-Data-Science/master/datasets/tea.csv> where the column separator is a comma and the first row includes the column names).

```{r}
tea <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/Helsinki-Open-Data-Science/master/datasets/tea.csv", sep = ",", header = T) # doesn't "first row includes the column names" mean header = F tho? 
```

> Explore the data briefly: look at the structure and the dimensions of the data. Use View(tea) to browse its contents. As you see, all variables are categorical. Convert them explicitly to factors, for example: tea\$sugar \<- factor(tea\$sugar) and then visualize the data.

```{r fig.width=10, fig.height=20}
str(tea) # before conversion
tea[sapply(tea, is.character)] <- lapply(tea[sapply(tea, is.character)], as.factor) # converting characters to factors
str(tea) # after conversion

library(ggplot2)
tea %>% dplyr::select(-age) %>% # removing age because it's an integer
  pivot_longer(cols = everything()) %>% 
  ggplot(aes(value)) + 
  facet_wrap("name", scales = "free") + 
  geom_bar() +
  labs(title = "Frequency of the variables") + # naming the plot
  theme(axis.text.x = element_text(size = 11, angle = 45, hjust = 1),
          # adjusting variable label size, angle and placement
        strip.text = element_text(size = 11, face = "bold"), 
          # adjusting titles of the plots (strips)
        plot.title = element_text(size = 22)) 
          # adjusting title size
```

### Multiple Correspondence Analysis

> Use Multiple Correspondence Analysis (MCA) on the tea data (or on just certain columns of the data, it is up to you!). Interpret the results of the MCA and draw at least the variable biplot of the analysis. You can also explore other plotting options for MCA. Comment on the output of the plots (0-4 points)

```{r}
library(FactoMineR)
mca <- tea %>% select(-age) %>% MCA

summary(mca)

mca %>% plot(
  invisible = c("ind"),
  habillage = "quali",
  graph.type = "classic"
)
```
