# Logistic Regression

### The data

```{r}
library(tidyverse); library(ggplot2); library(readr)
alc <- read_csv("C:/Users/SM/Documents/IODS-project/data/alc.csv")
colnames(alc)
```

The data set in use was contains demographic attributes and exam scores in Math and Portuguese of students in two Portuguese schools. You can view the variables above and read more about the data [here](https://archive.ics.uci.edu/ml/datasets/Student+Performance).

The purpose of this analysis is to use this data to study the relationships of high and low consumption of alcohol with sex, age, freetime and address. Before the analysis, I present the following hypotheses about these relationships:

-   sex: I expect more males to have high use of alcohol than females.

-   age: I expect high alcohol use to peak at 18 years old as it's the minimum age for legally selling alcohol outside restaurants in Portugal.

-   freetime: I expect those with more free time to be more likely to have a high alcohol consumption.

-   address: I expect more student living in urban areas to have a high alcohol use than those living in rural areas.

These hypotheses are not based on any data.

### Distributions of variables

Here I'll visualize the distribution of the variables and their relationship to alcohol use using both graphical and numerical means.

```{r}
# distribution of age
g_age <- alc %>% ggplot(aes(x = age))
g_age + geom_bar()
```

Just based on the age distribution alone my hypothesis is less plausible to be shown by this data as there is very little students above the age of 18 in this dataset.

```{r}
# distribution of sex
g_sex <- alc %>% ggplot(aes(x = sex))
g_sex + geom_bar()
```

```{r}
# distribution of address
g_address <- alc %>% ggplot(aes(x = address))
g_address + geom_bar()
```

Most of the students live in urban areas.

```{r}
# distribution of free time
g_freetime <- alc %>% ggplot(aes(x = freetime))
g_freetime + geom_bar()
```

The amount of experienced free time is approximately normal, though skewed to the right.

```{r}
# distribution of high alcohol use
g_high <- ggplot(data = alc, aes(x = high_use))
g_high + geom_bar()
```

Quite a lot of students have high alcohol consumption with the used criteria.

```{r}
# plotting alc_use and sex
gas <- ggplot(alc, aes(x = alc_use, y = sex))
gas + geom_boxplot()
```

```{r}
# plotting a plot of high_use and age
gha <- ggplot(alc, aes(x = high_use, y = age))
gha + geom_boxplot()
```

```{r}
# plotting alc_use and address
gaa <- ggplot(alc, aes(x = alc_use, y = address))
gaa + geom_boxplot()
```

Next, I'll show numerical summaries split into two tables to limit the number of rows:

```{r}
# produce summary statistics by group
alc %>% group_by(address, high_use) %>% summarise(count = n(), age = mean(age), freetime = mean(freetime))
```

```{r}
# produce summary statistics by group
alc %>% group_by(sex, high_use) %>% summarise(count = n(), age = mean(age), freetime = mean(freetime))
```

Based on these graphs and the summary, those with high alcohol consumption seem to have a little more free time, as I assumed. Those with high alcohol consumption seem a bit older but both groups are below 17 years old on average, contrary to my expectations. Higher proportion of students who live in a rural area have high alcohol consumption compared to those who live in an urban area, contrary to my assumptions. A higher proportion of males have a high alcohol use in comparison to females, as I assumed. As a side note, I find it interesting that females seem to have less free time than males, although the difference isn't large.

### Logistic regression model

Next, I'll explore the relationship between the presented variables using logistic regression.

```{r}
     
# creating a logistic regression model
m <- glm(high_use ~ sex + freetime + age + address, data = alc, family = "binomial")

# printing a summary of the model
summary(m)
```

From this summary I interpret that male sex, higher free time, higher age and living in a rural area are associated with high alcohol consumption. However, only sex and free time are statistically significant explanatory variables for high/low alcohol consumption.

```{r}
# compute odds ratios (OR)
OR <- coef(m) %>% exp

# compute confidence intervals (CI)
CI <- confint(m) %>% exp

# print out the odds ratios with their confidence intervals
cbind(OR, CI)
```

The odds ratio means that e.g. a male is about 2.2 times as likely to have a high use of alcohol. Each unit of free time (1-5) increases the odds of high use by 34 %, each year of age by 19 %. Those living in an urban area are about 0.65 times as likely to have high use of alcohol compared to those living in a rural area. However, the confidence intervals are wide and in the statistically insignificant variables *age* and *address* it includes 1, so the odds ratio values should not be taken literally by any means!

When compared to my hypotheses, the regression model confirms my expectations in terms of sex and free time, but contradicts my expectations with age and address. I don't repeat these in further details as these results align with the observations from the previous section.

### Predictive power of the model

```{r}
# predict() the probability of high_use
probabilities <- predict(m, type = "response")

alc <- alc %>% 
  mutate(probability = probabilities) %>% 
  mutate(prediction = probability > .5)

# tabulate the target variable versus the predictions
table(high_use = alc$high_use, prediction = alc$prediction)
```

From the table we can see that the prediction would be correct 244 + 18 = 262 times and mistaken 15 + 93 = 108 times.

```{r}
# initialize a plot of 'high_use' versus 'probability' in 'alc'
g <- ggplot(alc, aes(x = probability, y = high_use))

# define the geom as points and draw the plot
g + geom_point(aes(col = prediction))
```

```{r}
table(high_use = alc$high_use, prediction = alc$prediction) %>% prop.table %>% addmargins

```

In this table we can see the predictions as proportions. For example, using the given variables it can be correctly predicted about 66 % of the time that a student doesn't have a high alcohol consumption, and about 4 percent of time high consumption will be predicted when the alcohol consumption is not high.

And finally, calculating the training error:

```{r}
# define a loss function (mean prediction error)
loss_func <- function(class, prob) {
  n_wrong <- abs(class - prob) > 0.5
  mean(n_wrong)
}

loss_func(class = alc$high_use, prob = alc$probability)
```

This means that about 29 % of the predictions made using this model will be mistaken. This is better than random chance, where half the predictions would be mistaken, but neither is this impressive.
